A Global Picture of Food Insecurity and Hunger

### 2020-09-16 docs/modeling_fies.tex
Updates from Overleaf
model plots for paper, out-of-sample script

Given the subnational FIES data resulting from the disaggregation combined with the covariates stated above (See Table \ref{tab:covars}), we modelled subnational rates of severe and moderate-to-severe food insecurity for the whole world in the period of 2010 to 2030. We conduced suitable statistical methods such as Least Absolute Shrinkage and Selection Operator (LASSO) regression and Bayesian Stochastic Search and Variable Selection (Bayes SSVS), as well as machine learning methods such as Random Forest regression. Based on various model validation metrics, Random Forest (RF) regression performs best which can be seen as a contribution to the increasing importance of such methods in the field (CITATION: Master Thesis M\"uller?).

By using the most popular model validation metrics, we observed a in-sample mean squared error (MSE) of 0.00?? with a R2 of 0.999? for the RF regression model. Since these models are prone to overfitting (CITATION) it is crucial to check the out-of-sample prediction performance. Hence, we trained the RF regression model with 75\% of the data and predicted rates of severe and moderate-to-severe food insecurity for the remaining 25\%. We got an out-of-sample MSE of 0.00?? with a R2 of 0.99??. This excellent out-of-sample performance makes the RF regression model particularly suitable for our application. See the Supplement Info for a detailed description and validation of all models.

In RF regression models two main hyperparameters can be selected. One parameter which describes the number of variables randomly selected as candidates for splitting a node and another which describes the forest average number of unique cases (data points) in a terminal node, “mtry” and “nodesize” respectively (CITATION, manual r package randomforestSRC). We calculated the Out-Of-Bag (OOB) error for different combinations of the hyperparameters. The OOB error is the mean prediction error of the observation not included in the bootstrap aggregating or bagging (CITATION). By choosing the combination of hyperparameters with the lowest OOB error we found the optimal values which we used in our RF regression model to predict subnational rates of severe and moderate-to-severe food insecurity.

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{img/model/residuals.png}
  \caption{In-Sample Residuals of the RF regression model. Panel (A) shows the Moderate-to-Severe model and Panel (B) shows the Severe model.}
\end{figure}

>>>>

Given the subnational FIES data resulting from the disaggregation combined with the covariates stated above (See Table \ref{tab:covars}), we modelled subnational rates of severe and moderate-to-severe food insecurity for the whole world in the period of 2010 to 2030. We conduced suitable statistical methods such as Least Absolute Shrinkage and Selection Operator (LASSO) regression and Bayesian Stochastic Search and Variable Selection (Bayes SSVS), as well as machine learning methods such as Random Forest regression. Based on various model validation metrics, Random Forest (RF) regression performs best which can be seen as a contribution to the increasing importance of such methods in the field. For a detailed description and comparison of all models check M\"uller (CITATION: Master Thesis M\"uller 2020).

By using the most popular model validation metrics, we observed for the severe as well as for the moderate-to-severe RF regression model a in-sample mean squared error (MSE) of around zero with a R-squared (${R}^2$) of 0.99 (See Fig. \ref{fig:rf_in-sample}). Although RF regression models avoid overfitting by randomly selecting features at every node split (CITATION: Breiman 2001), it is still crucial to check the out-of-sample prediction performance. Hence, we trained the RF regression model with 75\% of the data and predicted rates of severe and moderate-to-severe food insecurity for the remaining 25\%. Similarly to the in-sample MSE and ${R}^2$, the out-of-sample MSE is close to zero with a ${R}^2$ of around 0.99 for both models (See Fig. \ref{fig:rf_out-sample}). This excellent out-of-sample performance makes the RF regression model particularly suitable for our application. 
%See the Supplement Info for a detailed description and validation of all models.

In RF regression models two main hyperparameters can be selected. One parameter which describes the number of variables randomly selected as candidates for splitting a node and another which describes the forest average number of unique cases (data points) in a terminal node, “mtry” and “nodesize” respectively (CITATION: Manual R-package randomforestSRC). We calculated the Out-Of-Bag (OOB) error for different combinations of the hyperparameters for the severe and moderate-to-severe model. The OOB error is the mean prediction error of the observations not included in the bootstrap aggregating or bagging (CITATION: Some fundamental work on Random Forest). By choosing the combination of hyperparameters with the lowest OOB error we found the optimal values for both model respectively.

Apart from tuning the hyperparameters, we examined the importance of the individual covariates (See Table \ref{tab:covars}) in explaining rates of severe and moderate-to-severe food insecurity. This is entirely based on the prediction error. By training the RF regression models with all covariates and predicting an outcome with and without a specific variables we calculated the average prediction error when leaving out that variable. According to the highest average prediction error we rank the importance of the models. We found that for both, the severe and moderate-to-severe model, the Poverty Headcount Index, Stunting, GDP Per Capita and the Gini Coefficient are the most important variables in explaining rates of severe and moderate-to-severe food insecurity (See Fig. \ref{fig:rf_vimp}).

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{img/model/in-sample_rf.png}
  \caption{In-Sample Fit of the RF regression model. Panel (A) shows the Moderate-to-Severe model and Panel (B) shows the Severe model.}
  \label{fig:rf_in-sample}
\end{figure}

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{img/model/out-sample_rf.png}
  \caption{Out-Of-Sample Fit of the RF regression model with 75\% training set and 25\% test set. Panel (A) shows the Moderate-to-Severe model and Panel (B) shows the Severe model.}
  \label{fig:rf_out-sample}
\end{figure}

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{img/model/vimp_rf.png}
  \caption{Variable Importance of the RF regression model. Panel (A) shows the Moderate-to-Severe model and Panel (B) shows the Severe model.}
  \label{fig:rf_vimp}
\end{figure}

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{img/model/error_rf.png}
  \caption{Error Rate of the RF regression model. Panel (A) shows the Moderate-to-Severe model and Panel (B) shows the Severe model.}
  \label{fig:rf_error}
\end{figure}
